Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU

### 2016-01-24 paper/paperSource.tex
paper changes...

% paper title
\title{Parallelization of Push-based System for Molecular Simulation Data Analysis}

@@

Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems generally follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on random I/O while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too.

Thankfully, there is a relatively new approach to this problem -- Push-based System for Molecular Simulation Data Analysis for processing network of queries proposed in the previous paper and it's advantages are: i) it uses traditional scan-based I/O framework to load the data from files to the main memory and then ii) the data is pushed through a network of queries which consequently filter the data and collect all the needed information. It has a considerable advantage in analysis of molecular simulation data, because it normally involves all the data sets to be processed by the queries.

In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster, but that might involve loss in data and some expectations from the type of data, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not needed in order to make it faster and still keep precision.

@@

\begin{IEEEkeywords}
Push-based system, molecular simulation, scientific databases,
spatial distance histogram, GPU, parallel processing.
\end{IEEEkeywords}
%\vspace{-5mm}

\begin{comment}
Many scientific applications share several features that bring in
light the shortcomings of the current data analysis systems. First,
the scientific analysis often involves executing a number of
analytical queries commonly processed on a large portion, if not all
of the generated data. Second, many of these fields use the same
popular, and small number of data analytics primitives as the basic
blocks on which they build their discovery. And three, the
scientific data, once stored is never modified. New data only
appends to the existing dataset, making the data perfect contender
for streaming.

And while the use of an index-based scan might seem better option to
a sequential scan, this is the case only when the query accesses
small portion of the data.
\end{comment}

@@

In various sciences simulation systems take big place and often times they may be very important for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations generally have the following flow \cite{mainPaper}:

@@

{\small
\begin{enumerate}
	\item initial physical properties are given to simulation software as arguments which are interpreted and provided in a language defined by the software
	\item then it runs for given amount of time that can generally last from milliseconds to months
	\item finally, the file generated during the simulation is analyzed to extract useful information
\end{enumerate}
}

\vspace{3mm}

The amount of data generated by scientific applications running on
advanced computer simulation systems as well as the new improved
experimental devices is growing rapidly. This big volume of data is
imposing a significant stress on the data analysis software. Even
though the existing data analysis systems are configured to deal
with large volume of data, they are not optimized for high
throughput data analysis. Often times, the data is available only
for certain period of time (e.g., streams) so the analyzing system
should react to it in a very efficient way. On top of the big amount
of data, another challenge that is being imposed to the data
analysis systems is that the data is usually being accessed through
high-level analytical queries which computation is a lot more
complex than the computation of simple aggregates. Some of these
queries are usually the bottleneck of the data analysis systems,
because they take a lot of time (often many days) to be executed and
the systems are not designed to handle multiple queries on the same
data stream at the same time, thus decreasing the overall efficiency
of the data analysis.

One such example, where fast data analysis is crucial involves
online social
media~\cite{nature_bigdata08,nature_bigdata12,science_social10}. The
social media is known to be used for real-time feedback collection,
which is usually associated to a particular topic or
product~\cite{jcs_twitter11}. In order for a system to be able to
perform analytical examination of the data produced in such
streaming media, the system should have the capability of fast data
access. The reason, the millions of data records(tweets) produced
every second~\cite{tkde_big14}. Moreover, these tweets may have
different geographical origin, introducing different languages and
forms and often times containing unsolicited messages, errors,
malicious content, etc. Therefore, some low level data uniformity
and cleaning on top of the data access and management issues should
be considered and possibly incorporated in the process of analytical
investigation in order to achieve relevant result.

The aforementioned big data imposed issues are also frequently
present in many other application fields, e.g., scientific data
analysis. In many scientific and engineering fields, the main
computational method for analyzing physical/chemical attributes of
natural systems is the particle simulation. Such simulations, when
performed in the field of structural and molecular biology, are
commonly called Molecular Simulations. Molecular simulations (MS)
(sometimes called Molecular Dynamics (MD) simulations) are computer
simulations of complex biological, physical or chemical structures.
These simulations, based on some basic theoretical models are widely
utilized as a basic research mechanism for studying the behavior of
the natural systems. The basic units of such systems/simulations are
natural particles (such as atoms, molecules, stars, etc.) and they
interact among each other for a certain period of time following
postulated classical forces. The number of these basic units taking
part in the simulations is big, usually in the range of hundreds of
thousands to millions. For instance, Fig.~\ref{fg:snap-white} shows
a snapshot of a collagen fiber, consisting of $890,000$ particles.
And oftentimes, the simulations produce datasets consisting of more
than one snapshot (frame) of the system's current state at various
time instants. Each of these frames contain all the particles,
together with all their measurements, such as spatial coordinates,
mass, charge, velocity, forces, etc. And usually, a big number of
such frames (in the tens of thousands) are being produced and stored
in the course of a typical simulation. Quantities measured during
the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale.

@@

\bibliographystyle{IEEEtran}
\bibliography{references}

>>>>

% paper title
\title{Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU}

@@

Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on random I/O while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too.

There is a new approach to this problem presented in the previous paper -- Push-based System for Molecular Simulation Data Analysis for processing network of queries proposed in the previous paper and its primary steps are: i) it uses traditional scan-based I/O framework to load the data from files to the main memory and then ii) the data is pushed through a network of queries which consequently filter the data and collect all the needed information which increases efficiency and data throughput. It has a considerable advantage in analysis of molecular simulation data, because it normally involves all the data sets to be processed by the queries.

In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. Using the same scan-based I/O framework the data is pushed through the network of queries which are processed by GPU, and due to the nature of science simulation data, this gives a big advantage for processing it faster and easier. In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster and those involved loss of data and some expectations from the data nature, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not even needed much though it didn't bear any loss in precision and performance.

@@

\begin{IEEEkeywords}
Push-based system, molecular simulation, scientific databases,
spatial distance histogram, GPU, parallel processing, CUDA.
\end{IEEEkeywords}
%\vspace{-5mm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

{\small
\begin{enumerate}
	\item initial physical properties are given to simulation software as arguments which are interpreted and provided in a language defined by the simulation software
	\item then it runs for given amount of time that can generally last from milliseconds to months or even more
	\item finally, the file generated during the simulation is analyzed to extract useful information
\end{enumerate}
}
\vspace{3mm}
The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software.

Big data processing is becoming one of the key issues with the amount of data being generated by modern systems. Eventually, this data is required to be processed on the fly, thus, analysis software should be able to handle massive data in a very short period of time. When working with huge volume of data, there are non-trivial issues arise. For example, it can take days and weeks to analyze big enough data sets, because the data cannot be simply loaded into memory, since it can get up to terabytes, thus, it has extra overhead because of widely used random access disk I/O framework to read from disk chunks by chunks. Besides it, analysis of the data can get even more complicated, since going through certain parts of the data once would not be enough, which leads to low throughput and efficiency of loaded data. One such example, the data analysis approach has polynomial complexity just for reading the data in order to come up with result, and since the data can't simply be saved in memory, it raises the overhead of disk I/O, too. Pull-based architectures in data processing engines are inefficient, since having a set of specific queries, in order to compute them all, it is needed to fetch data, filter it, and apply needed formulas. It is inefficient, since for every query the same chunk of data needs to be pulled into the memory at least the number of queries times or more in case of more complex queries.

As it has already been mentioned in the previous paper \cite{mainPaper}, one of the modern issues of analyzing massive data on the fly is social networks. .
"In order for a system to be able to perform analytical examination of the data produced in such streaming media, the system should have the capability of fast data access. The reason, the millions of data records(tweets) produced
every second. Moreover, these tweets may have
different geographical origin, introducing different languages and
forms and often times containing unsolicited messages, errors,
malicious content, etc. Therefore, some low level data uniformity
and cleaning on top of the data access and management issues should
be considered and possibly incorporated in the process of analytical
investigation in order to achieve relevant result. "\cite{mainPaper, nature_bigdata08,nature_bigdata12,science_social10,jcs_twitter11}

The primary focus and problem in this paper is scientific data analysis. Particles simulation is one of the most popular methods of analyzing certain chemical reactions, physical processes, or other behavior of different materials.
Molecular simulations (Molecular Dynamics) are applied in different fields and represent a method of analyzing physical movements of particles, atoms, and molecules in a fixed space with a given period of time, apparently with a possibility of giving initial state for each item that is involved in the process and can affect the system. This system is an N-body simulation. The number of atoms in simulations vary in hundred of thousands, particularly, we may observe two simulation systems of a collagen fiber structure and dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system consisting of 890,000 and 402,400 atoms respectively on Figure \ref{fg:collagen_dipal}. Simulation data represents number of records of physical properties such as mass, charge, velocity, coordinates, and forces for each item aggregated as frames, where each frame represents a snapshot of time, placed with a fixed time interval which may also vary depending on the simulation itself and simulation precision requirement. "Quantities measured during the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale." \cite{mainPaper}

Scientist gives the properties to simulation software (for example, Gromacs), runs the simulations, and finally get the output file. The output file must then be analyzed to produce certain results which may help him come up with certain consensus on original theoretical model that resulted in the molecular simulation system\cite{Frenkel:api01}. Gromacs is simulation software tool that helps scientist to run the actual simulation. It is a molecular dynamics package primarily designed for biomolecular systems such as proteins and lipids. \cite{Gromacs-online}. Besides the fact that it helps to generate the output files for the simulations, apparently it also helps to analyze the data itself, but the original problem is that it is not as optimized as it can be in order to analyze the data. Gromacs follows approach of pull-pased design, which means that for any given query (e.g. total mass or total charge, which are very similar type of 1-body queries without sophisticated selection) it will pull data separately and generate addition overhead wasting disk read I/O in order to come up with the result.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
In order to achieve objective discovery and to explain the workings
of the scientific systems, scientist must analyze the data produced
by the MS. These analysis oftentimes comprise of computation of very
complex quantities that show statistical properties of the data.
Such queries are of great importance to scientist because they are
the basic assembly blocks for a series of critical quantities needed
to outline the scientific systems \cite{Frenkel:api01}. The methods
by which the data is being accessed as well as these queries are
executed can either increase or decrease the efficiency of an data
analysis system. The known data analysis systems (e.g., Gromacs
\cite{Gromacs-online}) uses pull based type design in which the data
is being fed to the queries only by demand. When a query needs the
data, it requests it, the data is pulled through the system and
delivered to the query. This type of design, we believe, introduces
two types of problematic issues: 1. overhead in I/O traffic by not
allowing for multiple queries to be executed on the same stream of
data at the same time, and 2. cpu/data latency incurred when a
request for the data is sent to the system (the time it takes from a
request to the time the query can actually use the data).

%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems: a collagen fiber structure with 890,000 atoms (top) and a dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system with 402,400 atoms (bottom) \cite{mainArticle} }
 \label{fg:collagen_dipal}
\end{figure}

@@

\bibliographystyle{IEEEtran}
\bibliography{references.bib}
